---
title: "Code for Data Science Assissment"
author: "Martin"
date: "2024-12-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Required Libraries
```{r}
library(readxl)
library(writexl)
library(dplyr)       
library(lubridate)   
library(summarytools)
library(rmarkdown)
library(tidyverse)
```

### Load the dataset in R programming
```{r}
data_traffic = read_excel("C:/Users/GK TECH/Downloads/03 Real-Time_Traffic_Incident_Reports.xlsx")
head(data_traffic)
```


## Data Validation 

### Check for missing values in all variables in the dataset
```{r}
missing_values <- data_traffic %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Missing_Count")
print(missing_values)

```

### Checking for the format of the variables in the dataset
```{r}
str(data_traffic)
```

### Convert Published Date and Status Date from character to Date format
```{r}
data_traffic <- data_traffic %>%
  mutate(
    Published_Date = as.Date(Published_Date, format = "%m/%d/%Y %I:%M:%S %p %z"),
    Status_Date = as.Date(Published_Date, format = "%m/%d/%Y %I:%M:%S %p %z"))
head(data_traffic)
```

### Check uniqueness of Traffic Report ID

```{r}
sum(duplicated(data_traffic$`Traffic Report ID`))

```


### Extract Year and Month from Published_Date to easy the analysis and aggregation
```{r}
data_traffic <- data_traffic %>%
  mutate(
    Published_Year = year(Published_Date),
    Published_Month = month(Published_Date))
head(data_traffic)
```


### Extract Year and Month from Status_Date to easy the analysis and aggregation
```{r}
data_traffic <- data_traffic %>%
  mutate(
    Status_Year = year(Status_Date),
    Status_Month = month(Status_Date))
head(data_traffic)
```

### Impute missing values in Status variable with "Unknown"
```{r}
data_traffic <- data_traffic %>%
  mutate(
    Status = ifelse(is.na(Status), "Unknown", as.character(Status)),
  )
head(data_traffic)
```

### Convert Issue Reported and Status variable into factors
```{r}
data_traffic_clean<- data_traffic %>%
  mutate(
    Issue_Reported = as.factor(Issue_Reported),
    Status = as.factor(Status))
head(data_traffic_clean)
```

### EDA
### Summary statistics for all variables 

```{r}
summary(data_traffic_clean)
```

### Traffic Incidents by Year 2017-2022

```{r}
count_by_year <- data_traffic_clean %>%
  group_by(Published_Year) %>% summarise(Count = n()) %>% arrange(desc(Count))  

ggplot(count_by_year, aes(x = reorder(Published_Year, -Count), y = Count, fill = Count)) +
  geom_bar(stat = "identity") + 
  geom_text(aes(label = Count), vjust = -0.5, color = "black") +
  labs(title = "Traffic Incidents by Year", x = "Year", y = "Incident Count") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank())

```


### The number of traffic incidents monthly each year. 

```{r}
count_by_month <- data_traffic_clean %>%
  group_by(Published_Year, Published_Month) %>% summarise(Count = n()) %>% arrange(Published_Year, Published_Month)

ggplot(count_by_month, aes(x = Published_Month, y = Count, group = Published_Year, color = factor(Published_Year))) +
  geom_line() + geom_point() + geom_text(aes(label = Count), 
            vjust = 1.5,  color = "black", size = 3, 
            position = position_nudge(y = 5)) +  
  labs(title = "Traffic Incidents by Month and Year", x = "Month", y = "Incident Count") +  
  scale_x_continuous(breaks = 1:12, labels = month.abb) + 
  scale_color_brewer(palette = "Set1") +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),  
    axis.title.x = element_text(size = 14, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),  
    plot.title = element_text(size = 16, face = "bold"),
    axis.ticks.x = element_blank(), 
    legend.title = element_blank(),  
    legend.position = "bottom" 
  )

```



### Traffic Incidents by Type (Issue Reported) accross years 2017-2022

```{r}
count_by_year_type <- data_traffic_clean %>%
  group_by(Published_Year, Issue_Reported) %>%
  summarise(Count = n()) %>% arrange(Published_Year, desc(Count)) %>% group_by(Published_Year) %>% slice_head(n = 5) 
total_counts <- data_traffic_clean %>% group_by(Issue_Reported) %>% summarise(Total_Count = n()) %>% arrange(desc(Total_Count))

count_by_year_type <- count_by_year_type %>%
  mutate(Issue_Reported = factor(Issue_Reported, levels = total_counts$Issue_Reported))

ggplot(count_by_year_type, aes(x = reorder(Issue_Reported, -Count), y = Count, fill = Count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Count), vjust = 1.5, color = "black", size = 4) + 
  labs(title = "Top 5 Issue Reported type with highest Traffic Incidents by  in Each Year", 
       x = "Incident Type", 
       y = "Count of Incidents") +  # Add axis labels
  scale_fill_gradient(low = "lightblue", high = "darkblue") +   
  facet_wrap(~ Published_Year, scales = "free_y") +  
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),  
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14, face = "bold"),  
    axis.title.y = element_text(size = 14, face = "bold"),  
    plot.title = element_text(size = 16, face = "bold"),  
    axis.ticks.x = element_blank(),  
    legend.position = "none", 
    plot.margin = margin(10, 10, 40, 10)  
  )

```





